Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-04-02 paper/manuscript.tex
Addressed comments about mean/median and number of mutations.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean \comment{John: or is it median?} mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into XXX \comment{John: please add number} mutations pre-dating and XXX \comment{John: please add number} mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}A here so that they look consistent.}
}
\end{figure*}

The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
We observe both a shift in the mean \comment{John: or is it median?} mutation effect and in the distribution of mutation effects as we increase in circulation category.
Mutations that we measure to be substantially deleterious almost never reach high frequency.
Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.

>>>>

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
The empirical $P$-value represents the results of 1,000 permutations.
}
\end{figure*}

The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
We observe both a shift in the mean mutation effect and in the distribution of mutation effects as we increase in circulation category.
Mutations that we measure to be substantially deleterious almost never reach high frequency.
Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
